Physcraper: Body

### 2020-08-13 2-body.Rmd
a round to intro and methods

Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences. These sequences were gauged to have the potential to resolve phylogenetic relationships of most of the 241, 000 distinct taxa in the NCBI taxonomy at the time [about `r round(236023/240708, digits=4)*100`% of taxa; @sanderson2008phylota]. Assembling a DNA alignment from such a massive database can be done "by hand", but it requires huge amounts of time and it is mostly a non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to study evolutionary relationships among a variety of organisms [e.g., @smith2009mega; @antonelli2017toward; @izquierdo2014pumper].
<!-- *maybe a box of all pipelines?* -->
However, building molecular character matrices of few, thoughtfully curated markers and alignments can improve phylogenetic reconstructions [@fragoso2017pilot].

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a source of valuable expert knowledge. Linking published alignments with data from molecular databases that has not yet been included in any public phylogenetic estimate, has the potential to enrich and update phylogenetic relationships in many regions of the tree of life. The phylogenies associated to TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.
<!-- However, if there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, Phylesystem's metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link. -->

Physcraper relies on programmatic access protocols to automatically link molecular data from GenBank to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to continually update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for custom comparison of phylogenetic hypotheses.
Physcraper is coded as a Python pipeline that focuses on database interoperability, by integrating taxonomic name matching across biological databases. This integration makes downstream analyses straightforward, and allows users to place new data from a range biological database in an evolutionary context.

@@

```{r framework, echo=FALSE, fig.cap="The Physcraper software is fully described on its documentation website at physcraper.readthedocs.io, along with installation instructions, tutorials, examples and function usage documentation.", out.width = '85%', fig.retina= 2, fig.align = "center"}
# knitr::include_graphics("https://raw.githubusercontent.com/McTavishLab/physcraper/pyopensci/docs/img/schematic.svg")
knitr::include_graphics("docs/figs/schematic.pdf")
```

The general Physcraper framework is shown in Figure \@ref(fig:framework). Next, we will describe the technical details of each step of the workflow.

@@

- In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem]. If the user is not ready to make the input tree public, tree tip labels must be standardized to the unified OpenTree taxonomy using the bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool. *Should we say that it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?*
  This step is referred to as taxon name mapping. If taxonomic names can't be mapped, their taxonomic information will be excluded from further analysis.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the taxonomic name mapping and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline. Currently, only trees connected to a published study can be stored in the Phylesystem.
  <!-- Discussion: (although there are plans to allow storage of unpublished trees). -->
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on
  TreeBASE.
  <!--
  **EJM: Currently 2950 studies have treebase data links. It is possible that some/many of these have multiple trees in them, but that is harder to count.** -->
  <!-- there are `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website
  ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)).   -->

- The input alignment should be a single locus alignment that was used to generate the tree. Alignments are often stored in a public repository such as TreeBase [@piel2009treebase; @vos2012nexml],
  DRYAD ([www.datadryad.org](http://datadryad.org/)), or a data repository associated with the journal where the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper downloads it directly,
  either from the TreeBASE website ([www.treebase.org](https://treebase.org/))
  or through the TreeBASE GitHub repository (SuperTreeBASE; [github.com/TreeBASE/supertreebase](https://github.com/TreeBASE/supertreebase)).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
- Single locus alignments sometimes have fewer taxa than the tree inferred from the full concatenated data, simply because a single molecular marker usually does not cover all the taxa sampled for the full phylogenetic analysis. Physcraper prunes the input tree to taxa found in the alignment, and verifies that all taxon names on the tips of the tree are in the DNA character matrix and vice versa. Technically, just one taxon name (and its corresponding sequence in the alignment) is needed to continue the algorithm. See next section.
- A summary "csv" file with results from the taxon name standardization and pruning is produced for the user.
- The standardized and pruned tree and alignment (checked tree and alignment from now on) are output as "newick" and "fasta" respectively in the "inputs" folder to be used in the following steps.

@@

- Physcraper uses the GenBank DNA database as source to search for new sequences. The DNA sequence search can be performed on the GenBank remote database or in a GenBank local database set up by the user, which can speed up the search process. Detailed instructions to setup a local database are provided on Physcraper's software documentation.
<!-- Alternative DNA molecular databases could be used for this purpose, either personal or public databases, such as the European Nucleotide Archive or the DNA DataBank of Japan. The code to do this with Physcraper is not yet implemented. -->

- The next step is to identify a "search taxon" to constrain the sequence search on the GenBank database within that taxonomic group.
  The search taxon can be chosen by the user from the NCBI taxonomy.
  If none is provided, then the search taxon is automatically set using the taxa in the input tree labeled as the "ingroup" (Fig. \@ref(fig:framework)).
  The search taxon is The Most Recent Common Ancestor (MRCA) of the ingroup taxa in the OpenTree synthetic tree, that is also a named clade in the NCBI taxonomy.
  <!-- FIGURE RECOMMENDED: Figure \@ref(fig:search)  -->
  This is known in the OpenTree as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred as the Least Inclusive Common Ancestral taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.
  To identify the MRCAT of a group of taxon names, we use the OpenTree [taxonomic tool v3](<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>) [@rees2017automated].

@@

## Tree reconstruction and comparison

@@

# Examples

We will illustrate the utility of Physcraper in here with two use-case scenarios. One in which the user is interested in a particular group. Another one in which the user is interested in a particular phylogeny.
A tutorial as well as illustrated examples of commands for every step needed to perform a Physcraper analysis are available in the documentation website.

## The hollies

A student is interested in the genus *Ilex*, the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants.
The genus encompasses between 400-600 living species. A review of literature (google scholar search for "ilex phylogeny") reveals that there are several published phylogenetic trees showing relationships within the hollies [@cuenoud2000molecular; @manen2010history; @setoguchi2000intersectional; @selbach2009new], but only two have their data available publicly [@gottlieb2005molecular; @yao2020phylogeny].
@gottlieb2005molecular made tree and alignment data available in TreeBASE. The tree sampling 48 species was integrated to the OpenTree Phylesystem and is part of OpenTree's synthetic tree.
<!-- The second tree has not been made available anywhere, not even in the supplementary data of the original publication. -->
The most recent *Ilex* tree from @yao2020phylogeny has been made available in the OpenTree Phylesystem and in the DRYAD repository. It is the best sampled yet for the genus, with 200 species. However, it has not been added to OpenTree's synthtetic tree yet.
This makes it a perfect case to test the basic functionalities of Physcraper: we know that the sequences of the most recently published tree have been made available on the GenBank database. Hence, we expect that updating the oldest tree should at least contain the same species sampled in the largest tree.

>>>>

Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences. These sequences were gauged to have the potential to resolve phylogenetic relationships of most (`r round(236023/240708, digits=4)*100`%) of the almost 241, 000 distinct taxa in the NCBI taxonomy at the time [@sanderson2008phylota]. Assembling a DNA alignment from such a massive database can be done "by hand", but it requires huge amounts of time and it is mostly a non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to study evolutionary relationships among a variety of organisms [e.g., @smith2009mega; @antonelli2017toward; @izquierdo2014pumper].
<!-- *maybe a box of all pipelines?* -->
However, even during phylogenomic reconstructions, thoughtfully curated markers and alignments can improve phylogenetic reconstructions [@fragoso2017pilot].

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a source of valuable expert knowledge. Linking published alignments with public molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life. The phylogenies associated to TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.
<!-- However, if there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, Phylesystem's metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link. -->

Physcraper relies on programmatic access protocols to automatically link molecular data from GenBank to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to continually update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for comparison of existing phylogenetic hypotheses with newly generated ones.
Physcraper is coded as a Python pipeline that focuses on data interoperability, by integrating taxonomic name matching across biological databases. This integration also allows users to rapidly place new data from a diverse range of biological databases in an evolutionary context, making a variety of downstream analyses straightforward.

@@

```{r framework, echo=FALSE, fig.cap="The Physcraper framework consists of 4 steps (see text). The software is fully described on its documentation website at physcraper.readthedocs.io, along with installation instructions, function usage descriptions, examples and tutorials.", out.width = '85%', fig.retina= 2, fig.align = "center"}
# knitr::include_graphics("https://raw.githubusercontent.com/McTavishLab/physcraper/pyopensci/docs/img/schematic.svg")
knitr::include_graphics("docs/figs/schematic.pdf")
```

The general Physcraper framework is depicted in Figure \@ref(fig:framework). Briefly, it consists of 4 steps: 1) identifying and processing a phylogenetic tree to update and its underlying alignment; 2) performing a constrained BLAST search of sequences from the original alignment on the GenBank DNA database, and filtering of newly found sequences; 3) profile-aligning new sequences that passed the filtering to the original alignment; 4) performing a phylogenetic analysis and comparing the updated tree to previous phylogenetic estimates in the focus group. Next, we will describe technical details for each step.

@@

- In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem].
  Currently, only trees connected to a published study can be stored in the Phylesystem.
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on TreeBASE.
  If the user is not ready to make the input tree public, tree tip labels must be first standardized to the unified OpenTree taxonomy using OpenTree's bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool. *Should we say that it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?*
  This step is referred to as taxonomic name mapping and Phylesystem stored trees are processed in this way upon submission.
  Physcraper saves a summary "csv" file with results from the taxon name standardization for advantage of the user.
  If taxon names can't be mapped, their taxonomic information can't be used in comparison analysis. They will still be used in the sequence search and phylogenetic reconstruction steps.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the standardized taxonomic names and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline. 
  <!-- Discussion: (although there are plans to allow storage of unpublished trees). -->

  <!--
  **EJM: Currently 2950 studies have treebase data links. It is possible that some/many of these have multiple trees in them, but that is harder to count.** -->
  <!-- there are `r tryCatch({rotl::tol_about()$num_source_trees}, error = function(e) {1216})` published trees supporting the resolved nodes of the synthetic tree in the OpenTree website
  ([See OpenTree's website about](https://tree.opentreeoflife.org/about/synthesis-release/v12.3)).   -->

- The input alignment should be a single locus alignment that was used to generate the tree. Alignments are often stored in a public repository such as TreeBase [@piel2009treebase; @vos2012nexml],
  DRYAD ([www.datadryad.org](http://datadryad.org/)), or a data repository associated with the journal where the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper downloads it directly,
  either from the TreeBASE website ([www.treebase.org](https://treebase.org/))
  or through the TreeBASE GitHub repository (SuperTreeBASE; [github.com/TreeBASE/supertreebase](https://github.com/TreeBASE/supertreebase)).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
- Single locus alignments sometimes have fewer taxa than the tree inferred from the full concatenated data, simply because a single molecular marker usually does not cover all the taxa sampled for the full phylogenetic analysis. Physcraper prunes the input tree to taxa found in the alignment, and verifies that all taxon names on the tips of the tree are in the DNA character matrix and vice versa. Technically, just one taxon name (and its corresponding sequence in the alignment) is needed to continue the algorithm. See next section.
- The standardized and pruned tree and alignment (checked tree and alignment from now on) are output as "newick" and "fasta" respectively in the "inputs" folder to be used in the following steps.

@@

- Physcraper uses the GenBank DNA database as source to search for new sequences. The DNA sequence search can be performed on the GenBank remote database or in a GenBank local database set up by the user, which can speed up the search process. Detailed instructions to setup a local database are provided on Physcraper's software documentation.
<!-- Alternative DNA molecular databases could be used for this purpose, either personal or public databases, such as the European Nucleotide Archive or the DNA DataBank of Japan. The code to do this with Physcraper is not yet implemented. -->
- The next step is to identify a "search taxon" to constrain the sequence search on the GenBank database within that taxonomic group.
  The search taxon can be chosen by the user from the NCBI taxonomy.
  If none is provided, then the search taxon is automatically set using the taxa in the input tree labeled as the "ingroup" (Fig. \@ref(fig:framework)).
  The search taxon is The Most Recent Common Ancestor (MRCA) of the ingroup taxa in the OpenTree synthetic tree, that is also a named clade in the NCBI taxonomy.
  <!-- FIGURE RECOMMENDED: Figure \@ref(fig:search)  -->
  This is known in the OpenTree as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred as the Least Inclusive Common Ancestral taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.
  To identify the MRCAT of a group of taxon names, we use the OpenTree [taxonomic tool v3](<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>) [@rees2017automated].

@@

## Tree reconstruction

@@

# Example: The hollies

We will illustrate the utility of Physcraper in here with one use-case scenario in which a user is interested in a particular group, e.g., the genus *Ilex*, encompassing between 400-600 living species, is the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants, commonly known as "hollies".
A tutorial as well as illustrated examples of commands implementation on every step of the Physcraper workflow are available in the documentation website.

A review of literature (google scholar search for "ilex phylogeny") reveals that there are several published phylogenetic trees showing relationships within the hollies [@cuenoud2000molecular; @manen2010history; @setoguchi2000intersectional; @selbach2009new], but only two have their data available publicly [@gottlieb2005molecular; @yao2020phylogeny].
@gottlieb2005molecular made tree and alignment data available in TreeBASE. The tree sampling 48 species was integrated to the OpenTree Phylesystem and is part of OpenTree's synthetic tree.
<!-- The second tree has not been made available anywhere, not even in the supplementary data of the original publication. -->
The most recent *Ilex* tree from @yao2020phylogeny has been made available in the OpenTree Phylesystem and in the DRYAD repository. It is the best sampled yet for the genus, with 200 species. However, it has not been added to OpenTree's synthtetic tree yet.
This makes it a perfect case to test the basic functionalities of Physcraper: we know that the sequences of the most recently published tree have been made available on the GenBank database. Hence, we expect that updating the oldest tree should at least contain the same species sampled in the largest tree.
